A hazards approach to the biometric analysis of infant mortality

### 2023-04-04 paper2.Rmd
fix errors and add on natural time scale

```{r vbmap, fig.cap="Västerbotten."}
knitr::include_graphics("Slides/myfigs/vb.png")
```

@@

Given that the exogeneous mortality follows an Exponential distribution on the log-cube scale, it follows that the cumulative hazards function on the natural time scale is

\begin{equation*}
H(t; \lambda) = \lambda \log^3(t + 1), \; t > 0,
\end{equation*}

and it follows that the hazard function $h$ is given by

\begin{equation*}
h(t; \lambda) = \frac{3 \log^2(t + 1)}{t + 1}, \; t > 0,
\end{equation*}

@@

```{r scaniamap, fig.cap = "Skåne."}
knitr::include_graphics("Slides/myfigs/scania.png")
```

@@

However, this data set is known to be of very high quality [@tbcl99]. Sex ratio at birth 
is as expected and the neonatal mortality is high for all social groups. We are therefore inclined to 
think that the biometric model as it is described here simply is inappropriate during the eighteenth century.
Specifically, the log-cube transformation, which fits well with 19th century data, seems
to be completely out of order in the eighteenth century, maybe due to the high infant mortality,
which in this dataset is 222 per thousand.

@@

<!--- Let us look at the estimated cumulative hazards, total, 
exogenous and endgenous, see Figure \@ref(fig:escu).

```{r escu, fig.cap = "Scania 1710-1800.", eval = FALSE}
#source("R/biometric.R")
#biometric(sca1, graph = TRUE)
post <- age.window(sca1, c(g(28), 365))
lambda <- with(post, sum(event) / sum(exit - enter))
x <- c(0, rs$risktimes)
y <- x * lambda
hazo <- with(rs, c(0, n.events / size))
hazo <- cumsum(hazo)
plot(x, hazo, type = "s", col = "blue")
lines(x, y)
z <- hazo - y
lines(x, z, col = "red")
```

--->

@@

If anything, the hazards method never performs worse than the B-P method.
But the real strength of the hazards method is that it fits naturally into general 
modern survival analysis with censored and truncated data, and also proportional 
hazards models with covariates. 
<!--
On the other hand, the whole business with the 
biometric analysis and the log-cube transform seems to be highly questionable. 
It works for data in approximately the same time period and similar environment, 
but beyond that, it is not generalizable at all. These methods should be used with 
great care, if at all.
-->
Apparently, it works for many places in the mid-twentieth century, but beyond that, it is not generalizable. In fact, also when post-neonatal mortality is following a straight line on the log-cube scale, mortality may well deviate during different parts of the year. This means that the methods should be used with great care and only as a starting point.

>>>>

```{r vbmap, fig.cap="Västerbotten.", out.width="6in"}
knitr::include_graphics("Slides/myfigs/vb.png")
```

@@

Given that the exogenous mortality follows an Exponential distribution on the log-cube scale, 
it follows that the cumulative hazards function on the natural time scale is

\begin{equation*}
H(t; \lambda) = \lambda \log^3(t + 1), \; t > 0.
\end{equation*}

Note the duality: $\text{log-cube transform} \Longleftrightarrow  \text{cumulative hazards function}$!

It follows that the hazard function $h$ is given by

\begin{equation*}
h(t; \lambda) = \lambda \frac{3 \log^2(t + 1)}{t + 1}, \; t > 0,
\end{equation*}

@@

```{r scaniamap, fig.cap = "Skåne.", out.width="6in"}
knitr::include_graphics("Slides/myfigs/scania.png")
```

@@

However, this data set is known to be of very high quality [@tbcl99]. Sex ratio at birth 
is as expected and the neonatal mortality is high for all social groups. We are therefore inclined to 
think that the biometric model as it is described here simply is inappropriate during the eighteenth century.
Specifically, the log-cube transformation, which fits well with 19th century data, seems
to be completely out of order in the eighteenth century, maybe due to the high infant mortality,
which in this data set is 222 per thousand.

@@

Let us look at what happens on the natural time scale and with the Gompertz distribution 
replacing the log-cube time transformation, see Figure \@ref(fig:escu).

```{r escu, fig.cap = "Skåne 1710--1800, natural time scale, Gompertz exogenous distribution."}
#source("R/biometric.R")
#biometric(sca1, graph = TRUE)
par(mfrow = c(1, 1), las = 1, lwd = 2, cex = 0.7)
fit.c <- coxreg(Surv(enter, exit, event) ~ 1, data = sca11)
hc <- hazards(fit.c)
x <- c(0, hc[[1]][, 1])
y <- c(0, cumsum(hc[[1]][, 2]))
post <- age.window(sca11, c(28, 365))
fit.g <- phreg(Surv(enter, exit, event) ~ 1, data = post, dist = "gompertz", param = "rate")
hg <- hazards(fit.g)
plot(x, y, type = "l", xlab = "Day", ylab = "Cumulative hazards", axes = FALSE)
axis(1, at = c(0, 28, 183, 365))
att <-  c(0, 0.041, 0.063, 0.104, 0.216, 0.278)
axis(2, at = att)
abline(h = att, lty = 3, lwd = 1)
abline(h = 0, lwd = 1) 
abline(v = 28, lty = 2, lwd = 1, col = "darkgreen")
abline(v = 0, lwd = 1)
box()
y1 <- Hgompertz(x, shape = exp(-6.4115), rate = -0.007017,  param = "rate")
lines(x, y1, col = "blue", lty = 2)
xp <- x[c(0, 31:320)]
yp <- y - y1
lines(x, yp, col = "red", lty = 4)
```

Quite okay, if not perfect,

@@

If anything, the hazards method never performs worse than the B-P method.
But the real strength of the hazards method is that it fits naturally into general 
modern survival analysis with censored and truncated data, and also proportional 
hazards models with covariates. 
<!--
On the other hand, the whole business with the 
biometric analysis and the log-cube transform seems to be highly questionable. 
It works for data in approximately the same time period and similar environment, 
but beyond that, it is not generalizable at all. These methods should be used with 
great care, if at all.
-->
Apparently, it works for many places in the mid-twentieth century, but beyond that, 
it is not generalisable. In fact, also when post-neonatal mortality is following a 
straight line on the log-cube scale, mortality may well deviate during different 
parts of the year. This means that the methods should be used with great care and 
only as a starting point.
